\toolnameend: A Tool for Visualizing Patterns of Spreadsheet Function Combinations

### 2016-04-26 doc/paper.tex
Delirium at 2:30am. Fixes #154, fixes #155, fixes #156, fixes #157, fixes #158, fixes #159 (hopefully)

	\begin{figure*}[t] \centering \includegraphics[scale=.6]{IFargslabel} \caption{A
			picture of \toolname displaying the common arguments to the IF function's first 
			parameter} \centering \label{fig:fullpic} \end{figure*}

@@

	\subsection{Walkthrough of the Tool} \label{sec:walkthrough}Before discussing
	the minutiae of tool's development, we will give a basic example of how
	to use the tool. For this, we will answer this question: \textit{ What kinds of
		functions do people use to define the condition in an IF function?} The full
	visualization that answers this is available in Figure~\ref{fig:fullpic};
	however, to understand the interaction, we will start from the beginning. \par

@@

	The visualization so far comprises two types of nodes: the circle
	\includegraphics{glossary-greenonly}, which represents a discrete function in
	the formula and is size according to its frequency relative to the number of
	times to root function appeared\footnote{The root node, by definition, will
		always be the largest node in a tree.}; and the numbered squares
	\includegraphics{glossary-blue}, which represent the positions of arguments
	within its parent function. From this, we can confirm that, of the times it was
	observed, IF can have at most three arguments passed into it, which corresponds
	with its specifications in the
	API\footnote{https://support.office.com/en-us/article/IF-function-69aed7c9-4e8a-4755-a9bc-aa8bbff73be2}. \par

@@

	For both sides of the equal sign, the operator has certain types of arguments
	which predominate over the others: on the left side is most often a reference
	to another cell; on the right, a string or number literal, which makes sense
	for the case of confirming a value in another cell before assigning this one.
	Furthermore, a tooltip accompanies each function node in the tree, providing a
	concrete example of a function that uses this structure and where it can be
	found. If this single instance is not enough, then the user also has the option
	to double-click the individual function node, which will open up a new tab with
	a table of many more examples from dataset.

	\subsection{Icon Glossary}

	From these images, we can provide a quick glossary of images within the
	visualization to aid interpretation. \begin{itemize} \item
		\includegraphics{glossary-green} - \textbf{Function nodes} are green circles
		labeled with the function name. Their size is determined by their relative
		frequency, scaled logarithmically, to the frequency of the root node in the
		tree. They may have any number of arguments, including zero, as determined by
		the API.

		\item \vspace{.25cm}\includegraphics{glossary-leaf} - \textbf{Value nodes} are
		white circles with a generic label. They represent elements of a formula which
		are which accept no arguments, like references (A1), ranges
		(A1:A10)\footnote{That is, only if both sides of a references are cell
			references. If either side is a function that returns a cell reference, like
			INDIRECT, then the colon should be a function node.}, numbers, boolean values,
		strings, and errors. Functions with zero arguments, like NOW, however, should
		still be considered function nodes.

@@

		\item  \vspace{.25cm} \includegraphics{glossary-solidline} - \textbf{Solid
			lines} represent a function's ownership of its various arguments. As such,
		they connect the function nodes to argument nodes.

		\item  \vspace{.25cm} \includegraphics{glossary-fadingline} - \textbf{Fading
			lines} represent the possibility of an argument position to have several
		different possible functions or values. As such, they connect argument nodes
		to function or value nodes.

@@

		\item  \vspace{.25cm} \includegraphics[scale=.75]{glossary-lightblue} -
		\textbf{Optional argument trees} are variations of trees that show only
		formulae with a specific number of arguments, allowing the user to control for
		the presence of optional arguments. By default, trees with dark blue nodes
		contain every type of formula without regard how many arguments they have --
		for example, single-argument ANDs could appear alongside 4-argument ANDs. See
		Figure \ref{fig:optional} to see it visualized. \end{itemize}

	\begin{figure}[h] \centering \includegraphics[width =
		.46\textwidth]{comparison} \includegraphics[]{comparison-ss}
		\caption{The contents of each type of tree.} \label{fig:optional} \end{figure}

	\subsection{Implementation Details} The final visualization is the product of
	two discrete processes: \begin{itemize} \item \textit{Collection}: Given a set
		of Excel sheets, the tool, written in Java, uses Apache
		POI\footnote{https://poi.apache.org/} to identify and iterate over every cell
		containing a valid formula. Afterward, it calls POI's formula parser to break
		the formula text into an ordered set of individual tokens, which the tool then
		parses into the tree-like form. When all formulae have been analyzed like this,
		it produces JSON files for each top-level function in the set.

@@

	\subsection{Design Decisions} \label{ssec:decisions} Early in the design, we
	chose the tree form for its inherent ability to represent the loose
	parent-children relationship of formulae as functions and their arguments,
	which could be yet more functions. By adapting this structure to accommodate a
	broad range of possibilities for nested functions, the branching factor depends
	alternatively on the number of arguments in a function and the number of
	possible functions observed as an argument in a function. \par

	Guided by the goals in Section \ref{goals}, we faced a number of decision points, a
	sample of which we describe below, before we arrived at the design shown above.
	\begin{itemize}

		%Can I incorporate these again?
		%			\begin{itemize}
		%
		%				\item [1] \textbf{Provide an interactive environment for free
		% exploration.}
		%				There are a lot of data in some of these collections, and the user should
		% be
		%				able to pick for themselves which to explore while ignoring the rest.
		% Furthermore,
		%				this exploration should not be designed around answering one specific
		% question
		%				but rather be open-ended.
		%
		%				\item [!1] \textbf{Do not make judgments about good and bad design.}
		% Though
		%				users may certainly approach the tool with the intention of finding
		% particular
		%				formulae, \toolname will not prioritize any on metrics of good or bad
		%				practice. Rather, it will present them agnostically, referring only to
		%				frequency.
		%
		%				\item [2] \textbf{Emphasize the quantitative patterns in formula
		%					construction.} \toolname should address the questions of how often the
		%				spreadsheet programmers used a certain function and where they used it.
		% In
		%				this way, it must show precise metrics, such as frequency of use and
		% depth
		% of
		%				function nesting, of the dataset it conveys.
		%
		%				\item [!2] \textbf{Do not overwhelm with numbers.} For popular functions
		%				especially, we anticipated some particularly dense trees. Though
		%				everything should be accessible in accordance with Goal 1, we should
		% design
		% in
		%				a way that reduces noise.
		%
		%				\item [3] \textbf{Promote a qualitative understanding of the patterns.}
		%				\toolname should supplement its observations with concrete instances of
		%				relevant formulae from the corpus. Ideally, it should even direct users
		% to
		% the
		%				exact cell in the spreadsheets where the formula was used,
		% contextualizing
		% the
		%				functions.
		%
		%				\item [!3] \textbf{Do not attempt to explain formulae.} Though \toolname
		% tries
		%				to foster understanding by linking pattern to example, it should not
		% provide
		%				a precise description of what a complex formula accomplishes. The tool's
		% user
		%				must infer this on their own.
		%
		%			\end{itemize}

>>>>

@@

	\subsection{Walkthrough of the Tool} \label{sec:walkthrough}Before discussing
	the minutiae of tool's development, we will give a basic example of how
	to use the tool. For this, we will answer this question: \textit{ What kinds of
		functions do people use to define the condition in an IF function?} \par

@@

	The visualization so far comprises two types of nodes: the circle
	\includegraphics{glossary-greenonly}, which represents a discrete function in
	the formula and is size according to its frequency relative to the number of
	times to root function appeared; and the numbered squares
	\includegraphics{glossary-blue}, which represent the positions of arguments
	within its parent function. From this, we can confirm that, of the times it was
	observed, IF can have at most three arguments passed into it, which corresponds
	with its specifications in the
	API\footnote{https://support.office.com/en-us/article/IF-function-69aed7c9-4e8a-4755-a9bc-aa8bbff73be2}. \par

@@

	\begin{figure*}[t] \centering \includegraphics[scale=.5]{IFargslabel} \caption{A
			picture of \toolname displaying the common arguments to the IF function's first 
			parameter} \centering \label{fig:fullpic} \end{figure*}

	For both sides of the equal sign, the operator has certain types of arguments
	which predominate over the others: on the left side is most often a reference
	to another cell; on the right, a string or number literal, which makes sense
	for the case of confirming a value in another cell before assigning this one.
	Furthermore, a tooltip accompanies each function node in the tree, providing a
	concrete example of a function that uses this structure. If this single instance is not enough, then the user also has the option
	to double-click the individual function node, which will open up a new tab with
	a table of many more examples from dataset.

	\subsection{Glyph Glossary}

	From these images, we can provide a quick glossary of glyphs within the
	visualization to aid interpretation. \begin{itemize} \item
		\includegraphics{glossary-green} - \textbf{Function nodes} are green circles
		labeled with the function name. Their size is determined by their relative
		frequency, scaled logarithmically, to the frequency of the root node in the
		tree. They may have any number of arguments, including zero, as determined by
		the API.

		\item \vspace{.25cm}\includegraphics{glossary-leaf} - \textbf{Value nodes} are
		white circles with a generic label. They represent elements of a formula which
		accept no arguments, like references (A1), ranges (A1:A10), numbers, boolean
		values, strings, and errors. However, \toolname treats still functions with
		zero arguments, like NOW, as function nodes.

@@

		\item  \vspace{.25cm} \includegraphics{glossary-solidline} - \textbf{Solid
			lines} represent a function's acceptance of its parameters. As such,
		they connect the function nodes to argument nodes.

		\item  \vspace{.25cm} \includegraphics{glossary-fadingline} - \textbf{Fading
			lines} represents how a formula author can use several different possible
		functions or values within a single argument position. As such, they connect
		argument nodes to function or value nodes.

@@

		\item  \vspace{.25cm} \includegraphics[scale=.75]{glossary-lightblue} -
		\textbf{Optional argument trees} are variations of trees that show only
		formulae with a specific number of arguments. For example, the function AND can
		take any number of arguments, up to 255. By default, the tree shows data from
		AND functions with any number of argument, but these trees, marked with a darker
		blue color, can show how AND was used when \textit{only} two arguments were passed into 
		it, or \textit{only} four, and so on. See
		Figure \ref{fig:optional} for how the data between tree types differs. \end{itemize}

	\begin{figure}[h] \centering \includegraphics[width =
		.46\textwidth]{comparison} \includegraphics[]{comparison-ss}
		\caption{The contents of the default tree (left) and the optional argument trees (right)} \label{fig:optional} \end{figure}

	\subsection{Implementation Details} The visualization is the product of
	two discrete processes: \begin{itemize} \item \textit{Collection}: Given a set
		of Excel sheets, the tool, written in Java, uses Apache
		POI\footnote{https://poi.apache.org/} to identify and iterate over every cell
		containing a valid formula. Afterward, it calls POI's formula parser to break
		the formula text into an ordered set of individual tokens, which the tool then
		parses into the tree-like form. When all formulae have been analyzed like this,
		it produces JSON files for each top-level function in the set.

@@

	\subsection{Design Decisions} \label{ssec:decisions} Early in the design, we
	chose the tree form for its inherent ability to represent the
	parent-children relationship of formulae as functions and their arguments,
	which could be yet more functions. By adapting this structure to accommodate a
	broad range of possibilities for nested functions, the branching factor depends
	alternatively on the number of arguments in a function and the number of
	possible functions observed as an argument in a function. \par

	Guided by the goals in Section \ref{goals}, we faced a number of decisions in how
	to represent the data in the visualization. Below, I've outlined those that had
	the largest impact in the values that appear in the tree.
	\begin{itemize}
